Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-06-12 large_number_squeezing_with_atoms.tex
coupling between spatial modes -> momentum modes
What was said previously was wrong.  It should be nonlinearity-induced coupling between momentum modes.
Added a brief summary of the Bog derivation.

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sinc}{sinc}
\newcommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\expect}[1]{\ensuremath{\left<#1\right>}}
\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}

@@

To find the trap geometries that produce the best squeezing, it is important to understand the root cause of this loss of squeezing in higher dimensional traps.  The dependence on dimensionality is not explained by the processes described in section \ref{sec:MMdescription}, which focus on the effects of inhomogeneity within the trap.  Fundamentally, this degradation of the squeezing is due to nonlinearity-induced coupling between different spatial modes.

@@

Here we put some details of the Bog analysis.

\begin{align}
  S &= \frac{\Var(n_0) + \sum_k \frac{1}{2} n_k (3 + \cos 2\theta + 4 n_k)}{n_0 + \sum_k n_k} \\
  n_k &= \left[\lambda \cos\theta \sinc(\omega_k \tau)\right]^2 \\
  \hat{a}(t_3) &= \left[\hat{a}(0) - i \chi \alpha \tau_h\left(\alpha^* \hat{a}(0) + \alpha \hat{a}^\dagger(0) - 2\abs{\alpha}^2\right) \right]\cos(\theta) - i\hat{b}(0) \sin\theta e^{i\phi} \\
  n_0 &= \left(\alpha ^2+\alpha ^4 \lambda ^2\right) \cos^2\theta + \left(\beta^2 \sin\theta  + 2\alpha\beta\cos\theta\sin\phi\right)\sin\theta\\
  \begin{split}
  \Var(n_0) &= \frac{1}{2} \left(\alpha ^2+\beta ^2\right)+\frac{1}{8}\alpha ^4 \left(7+ 2\beta ^2\right) \lambda ^2+ \frac{3}{4} \alpha ^8 \lambda ^4 \\
  &\phantom{\mathrel{=}}+\frac{1}{2}\left( \alpha ^2-\beta ^2+2 \alpha ^4 \lambda ^2+2 \alpha ^8 \lambda ^4\right) \text{Cos}[2 \theta ] \\
  &\phantom{\mathrel{=}}+\frac{1}{8}\left(\alpha ^4 \lambda ^2 \left(1-2 \beta ^2+2 \alpha ^4 \lambda ^2\right) \text{Cos}[4 \theta ] \right) \\
  &\phantom{\mathrel{=}}+\frac{1}{8}\left(4 \alpha  \beta  \left(8 \alpha ^2 \lambda  \text{Cos}[\theta ]^3 \text{Cos}[\phi ] \text{Sin}[\theta ]+\text{Sin}[2 \theta ] \left(2 \text{Sin}[\phi ]+\alpha  \beta  \lambda  \text{Sin}[2 \theta ] \left(\alpha ^2 \lambda  \text{Cos}[2 \phi ]+\text{Sin}[2 \phi ]\right)\right)\right)\right)
  \end{split}
\end{align}

>>>>

\DeclareMathOperator{\Var}{Var}
\DeclareMathOperator{\sinc}{sinc}
\newcommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\newcommand{\expect}[1]{\ensuremath{\left<#1\right>}}
\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}
\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}

@@

To find the trap geometries that produce the best squeezing, it is important to understand the root cause of this loss of squeezing in higher dimensional traps.  The dependence on dimensionality is not explained by the processes described in section \ref{sec:MMdescription}, which focus on the effects of inhomogeneity within the trap.  Fundamentally, this degradation of the squeezing is due to nonlinearity-induced coupling between different momentum modes.

@@

The Bogoliubov model presented in Section~\ref{sec:2D3Dsqueezing} is taken in the limit of a homogeneous condensate with $\chi_{aa}\neq 0$, $\chi_{ab}=\chi_{bb}=0$.  In this limit, the Hamiltonian for the internal state $\ket{a}$ during the hold time is
\begin{align}
  \hat{H}_{a} &= \int \left[ \hat{\psi}_a^\dagger \left(-\frac{\hbar^2}{2 m}\nabla^2\right) \hat{\psi}_a + \frac{U_{aa}}{2} \hat{\psi}_a^\dagger \hat{\psi}_a^\dagger \hat{\psi}_a \hat{\psi}_a \right] \,dV \label{eqBogOriginalHamiltonian}
\end{align}
To find the response of the condensate to small fluctuations we define fluctuation operators for the non-zero momentum modes $\delta\hat{a}_\mathbf{k} = \hat{a}_\mathbf{k} - \alpha_\mathbf{k}$ where $\alpha_\mathbf{k} = \expect{\hat{a}_\mathbf{k}}$ is the condensate mean field ($\alpha_\mathbf{k}=0$ for $\mathbf{k} \neq 0$ as the condensate is homogenous).  The Hamiltonian \eqref{eqBogOriginalHamiltonian} is then expanded in terms of $\delta\hat{a}_\mathbf{k}$ retaining only the lowest order contributions.  The lowest order contribution is second order:
\begin{align}
  \hat{H}_{a}^{(2)} &= \sum_{\mathbf{k}\neq 0} \left[ \frac{\hbar^2 \mathbf{k}^2}{2M} \delta\hat{a}_\mathbf{k}^\dagger \delta\hat{a}_\mathbf{k} + \frac{1}{2} \hbar \chi_{aa} n_a \left(\delta\hat{a}_\mathbf{k}^\dagger \delta\hat{a}_{-\mathbf{k}}^\dagger + \delta\hat{a}_{-\mathbf{k}} \delta\hat{a}_{\mathbf{k}} \right) \right].
\end{align}
This Hamiltonian is diagonalised with the standard Bogoliubov transformation \cite{PethickSmith}
\begin{align}
  \delta \hat{a}_\mathbf{k}(t) &= u_\mathbf{k} \hat{b}_\mathbf{k}(t) - v_k \hat{b}_{-\mathbf{k}}^\dagger(t),\\
  u_k^2 &= v_k^2+1 = \frac{1}{2} \left(\frac{\omega^0_\mathbf{k} + 2 \chi_{aa}n_a}{\omega_\mathbf{k}} +1\right),
\end{align}
where $\omega^0_\mathbf{k} = \hbar \mathbf{k}^2/2M$ is the frequency of a free particle, and $\omega_\mathbf{k} = \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}$ is the frequency of the Bogoliubov modes:
\begin{align}
  \delta\hat{b}_\mathbf{k}(t) &= \exp\left(-i \omega_\mathbf{k} t\right)\delta\hat{b}_\mathbf{k}(0).
\end{align}

The expectation value of the number of atoms in momentum mode $\hbar\mathbf{k}$ after the hold time is then found to be
\begin{align}
  \expect{\hat{n}_\mathbf{k}(t_2)} &= \expect{\delta\hat{a}^\dagger_\mathbf{k} \delta\hat{a}_\mathbf{k}} \\
  \begin{split}
  &=\left[1 + 4 u_\mathbf{k}^2(2 u_\mathbf{k}^2 - 1)\sin^2(\omega_\mathbf{k} \tau_\text{hold})\right] \expect{\delta \hat{a}^\dagger_\mathbf{k} \hat{a}_\mathbf{k}} \\
  &\phantom{\mathrel{=}}- 4 u_\mathbf{k} v_\mathbf{k} \sin(\omega_\mathbf{k}\tau_\text{hold}) \Im\left\{\left[1-i2 u_\mathbf{k}^2 \sin(\omega_k \tau_\text{hold})\right] \expect{\hat{a}_{-\mathbf{k}} \hat{a}_\mathbf{k}}\right\}\\
  &\phantom{\mathrel{=}} + 4 u_\mathbf{k}^2 v_\mathbf{k}^2 \sin^2(\omega_\mathbf{k}\tau_\text{hold})
  \end{split}
\end{align}
If the non-zero momentum modes are initially in a vacuum state then this reduces to
\begin{align}
  \expect{\hat{n}_\mathbf{k}(t_2)} &= 4 u_\mathbf{k}^2 v_\mathbf{k}^2 \sin^2(\omega_\mathbf{k}\tau_\text{hold}), \\
  &= \left[n_a \chi_{aa} \tau_\text{hold} \sinc(\omega_\mathbf{k} \tau_\text{hold}) \right]^2.
\end{align}
After the final Rabi coupling pulse the occupation of the non-zero momentum mode $\hbar\mathbf{k}$ is
\begin{align}
  n_a(\mathbf{k}, t_3) &= \left[n_a \chi_{aa} \tau_\text{hold} \cos(\theta)\sinc(\omega_\mathbf{k} \tau_\text{hold}) \right]^2,
\end{align}
which is the expression given earlier in Eq.~\eqref{eqBogOccupation}.
